\bf Adaption of the Chumbley Score to matching of bullet striation marks

### 2018-05-23 doc/paper.Rmd
conclusion + stab at abstract

---
title: "Adaption of the Chumbley Score to matching of bullet striation marks"
authors:
- affiliation: Department of Statistics, Iowa State University 
  name: Ganesh Krishnan
  thanks: The authors gratefully acknowledge ...
- affiliation: Department of Statistics and CSAFE, Iowa State University 
  name: Heike Hofmann
biblio-style: apsr
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- forensic science
- toolmark
- cross-correlation
- Mann-Whitney U statistic
- 
- 
bibliography: bibliography
abstract: null
---

@@

Analysis of these digitized markings require the use of statistical methods which can quantify the scientific mechanism of comparing markings and serve as basis for an error rate calcluation.
\citet{manytoolmarks2} define a relative distance metric and use it as similarity measure between two toolmarks. \citet{manytoolmarks1} extract many small segments in the markings of two toolmarks and compare similiarity using a maximum pearson correlation coefficient. 
The Chumbley scoring method, first introduced by \citet{chumbley}, uses a similar but more extensive framework based on a Mann-Whitney U test of the resulting correlation coefficients. This approach is non-deterministic, because segments are chosen randomly.  \citep{hadler} make the score deterministic for each pair of toolmarks by choosing segments for comparison systematically. This approach also ensures independence between segements of striae.  In this paper, we are investigating the applicability of the Chumbley scoring method by \citet{hadler} to assess striation marks on bullet lands for same-source identification.

@@

Striation marks on bullets are made by impurities in the barrel. As the bullet travels through the barrel, these imperfections leave "scratches" on the bullet surface (see top of \autoref{fig:rgl}). Typically, only striation marks in the land engraved areas (LEAs) are considered \citep{afte-article1992}. Bullet lands are depressed areas between the  grooves made by the rifling action of the barrel. Compared to toolmarks made by screwdrivers striation marks on bullets are typically much smaller, both in length and in width. Bullets also have a curved cross-sectional topography.

@@

<!--Bullet matching methods are usually based on these associated signatures.-->
In same-source comparisons this curvature is usually removed using some form of Gaussian filter \citep{ma2004} or non-parametric smoothing \citep{aoas}. 
\citet{chu2013} use an automatic method for counting consecutive matching striae (CMS). The authors report an error rate of 52% for known same-source land comparisons to be (incorrectly) identified as different-source (false negative) and zero false positives for known different-source lands. \citet{ma2004} and \citet{vorburger2011} discuss CCF (cross-correlation function) and its discriminating power and applicability for same-source analyses of bullets, but do not provide any error rates in their discussion. \citet{aoas} use multiple features, such as CCF, CMS, D (distance measure), etc. in a random forest based method and compare every land against every other land of  digitised versions of Hamby 252 and Hamby 44 \citep{hamby} published on the NIST Ballistics Database \citep{nist}. The authors report an out-of-bag overall error rate of 0.46%, comprised of an error rate of 30.05% of  same-source pairs that were not identified and an error rate of 0.026% of different-source pairs that were incorrectly identified as same-source.

@@

@@

In a pre-processing step  the two markings are smoothed using a lowess \citep{lowess} with coarseness parameter $c$. 
 Originally, this smoothing is intended to remove drift and (sub)class characteristics from individual markings, however, in the setting of matching bullet striae, we can also make use of this mechanism to separate bullet curvature in profiles from signatures before matching signatures.  \autoref{fig:sigs-profiles} shows an example of a bullet land profile (left) and the corresponding signature (right).

@@

\autoref{fig:eer} shows a comparison of the performance of the two methods CS1 and CS2 with respect to EER (eaual error rate) and AUC (area under the curve) corresponding to the ROC curves shown in \autoref{fig:roc}. Equal error rates are reduced using method CS2, while area under the curve significantly increases (at a significance level $\alpha$ of 0.05) compared to method CS1.

@@

cvcs %>% filter(!wo %in% c(130, 140)) %>%
  ggplot(aes(d = as.numeric(match), m = 1-p_value)) + 
  geom_roc(labels=FALSE, n.cuts=100, pointsize=0, size=0.6, 
           aes(colour = method.wo)) +
  facet_wrap(~wv, labeller = label_bquote(w[v] == .(wv))) +
  scale_alpha_discrete(range=c(0.1, 0.6)) +
  scale_colour_manual(expression("Method /\nOptimization\nwindow size"~w[o]), values=cols) +
  theme_bw() +
  xlab("False Positives Rate (1 - Specificity)") +
  ylab("True Positives Rate (Sensitivity)") +
  guides(
   colour = guide_legend(ncol=2)
  ) +
  geom_abline(slope=-1, intercept = 1, linetype = 3, 
              colour="grey30", size = 0.25)

@@

\begin{table}[hbtp]
\centering
\begin{adjustbox}{max width=\textwidth}
\begin{tabular}{rrrlrrl}
  \hline
  & \multicolumn{3}{l}{\bf CS1} & \multicolumn{3}{l}{\bf CS2} \\
$w_o$ & \bf FPR & \bf FNR & \bf AUC (95\% C.I.) & \bf FPR & \bf FNR & \bf AUC (95\% C.I.) \\ 
  \hline
90 & 0.068 & 0.330 & 0.850 (0.835, 0.865) & 0.075 & 0.242 & 0.877 (0.863, 0.890) \\ 
  120 & 0.065 & 0.309 & 0.864 (0.850, 0.878) & 0.072 & 0.236 & 0.890 (0.877, 0.903) \\ 
  130 & 0.065 & 0.317 & 0.863 (0.850, 0.877) & 0.072 & 0.224 & 0.898 (0.886, 0.910) \\ 
  140 & 0.065 & 0.304 & 0.873 (0.859, 0.886) & 0.071 & 0.229 & 0.902 (0.891, 0.914) \\ 
  150 & 0.067 & 0.326 & 0.863 (0.850, 0.877) & 0.072 & 0.225 & 0.904 (0.892, 0.916) \\ 
  180 & 0.061 & 0.326 & 0.877 (0.864, 0.890) & 0.070 & 0.229 & 0.907 (0.896, 0.919) \\ 
  210 & 0.064 & 0.342 & 0.865 (0.851, 0.878) & 0.066 & 0.231 & 0.906 (0.895, 0.918) \\ 
   \hline
\end{tabular}
\end{adjustbox}
\caption{\label{tab:aucs} Overview of results as shown in Figures \ref{fig:roc} and \ref{fig:eer}. FPR is the observed rate of false positives (for a fixed level $\alpha = 0.05$) or type I error rate, FNR is the false negative rate or type II eror rate. Area under the curve (AUC) is shown with confidence intervals. }
\end{table}

@@

The high-resolution scans of Hamby set 44 are capturing images at a resolution of $0.645 \mu m$ per pixel. On average, land engraved areas are 3000 pixels in length. A coarseness of $c = 0.15$ was found to be sufficient in removing any bullet curvature.
\autoref{fig:h44} shows the resulting EER and AUC for methods CS1 and CS2 based on two sizes of validation windows ($w_v \in 75, 125$) and optimization window sizes around 300 pixels (10 percent of the average length). Interestingly, the overall performance of both CS1 and CS2 is a lot lower for the high resolution version of Hamby-44 than for the lower resolution scans. The area under the curve overall is significantly lower for the high resolution scans than for the previous set of scans. Partly, this might be due to the particular choice of the parameters.

@@

\hh{
We started out trying to assess the suitability of the Chumbley Score's for matching striae on bullet lands. Besides using recommended defaults, we also proposed ways to optimize parameters: based on the assumption that once sub-class characteristics are removed, optimal locations are distributed uniformly across the profile, we found  for bullets the smaller coarseness value of $c = 0.15$ to be suitable. 
Optimized window sizes were found based on cross-validation to minimize type 2 error rates. 
Method CS1 proposed by \citet{hadler} has a minimal type 2 error rate of XXX for an optimized window size of YYY - which is considerably higher than the error rates achieved on matching toolmarks. 
Unfortunately, method CS1 also has a high rate of failed tests -- situations, in which the algorithm does not provide a result, due to the way different-shift pairs are constructed. 
XXX
The algorithm CS2 introduced here is constructed in a way that reduces the number of failures by a magnitude of about XXX. While reducing the failure rate, the algorithm also shows an increase in the power of the test. Type II error rates reach a minimum of XXX for an optimized window size of YYY. While significantly reduced over CS1, CS2 still has type 2 error rates on bullet lands that are  higher than the error rates achieved on the -much larger- toolmarks.
However, bullets usually have multiple lands - in the case of Ruger P85s as used in the Hamby study, there are six lands for each bullet. We might be able to get more power out of the test by adapting CS2 to deal work in a bullet-to-bullet comparison.
}

\hh{
Fix in CS2 should also improve power for matching toolmarks. 
}

@@

\caption{\label{same-shift-failure}Sketch of same-shift pairings. When the lag between optimal reference points is too large to accommodate a validation window in noth signatures, both CS1 and CS2 fail.}
\end{figure}

>>>>

---
title: "Adaption of the Chumbley Score to matching of bullet striation marks"
authors:
- affiliation: Department of Statistics, Iowa State University 
  name: Ganesh Krishnan
  thanks: The authors gratefully acknowledge funding from CSAFE through cooperative agreement \# 70NANB15H176 between NIST and Iowa State University. 
- affiliation: Department of Statistics and CSAFE, Iowa State University 
  name: Heike Hofmann
biblio-style: apsr
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- forensic science
- toolmark
- cross-correlation
- Mann-Whitney U statistic
- 
- 
bibliography: bibliography
abstract: Same-source analyses are a major part of a Forensic Toolmark Examiner's job. Technological advances, such as profilometers and confocal microscopy allow to measure 3D surfaces in a high-resolution digitized form. Methods for matching marks for a variety of tools have been studied in the literature. Analysis of these digitized markings require the use of statistical methods which can quantify the scientific mechanism of comparing markings and serve as basis for an error rate calculation. In this paper, we are investigating the applicability of the Chumbley scoring method by \citet{hadler} to assess striation marks on bullet lands for same-source identification. The Chumbley score  provides us with another approach in the same-source assessment of bullet striation marks. 
---

@@

Analysis of these digitized markings require the use of statistical methods which can quantify the scientific mechanism of comparing markings and serve as basis for an error rate calculation.
\citet{manytoolmarks2} define a relative distance metric and use it as similarity measure between two toolmarks. \citet{manytoolmarks1} extract many small segments in the markings of two toolmarks and compare similarity using a maximum Pearson correlation coefficient. 
The Chumbley scoring method, first introduced by \citet{chumbley}, uses a similar but more extensive framework based on a Mann-Whitney U test of the resulting correlation coefficients. This approach is non-deterministic, because segments are chosen randomly.  \citep{hadler} make the score deterministic for each pair of toolmarks by choosing segments for comparison systematically. This approach also ensures independence between segments of striae.

@@

In this paper, we are investigating the applicability of the Chumbley scoring method by \citet{hadler} to assess striation marks on bullet lands for same-source identification.
Striation marks on bullets are made by impurities in the barrel. As the bullet travels through the barrel, these imperfections leave "scratches" on the bullet surface (see top of \autoref{fig:rgl}). Typically, only striation marks in the land engraved areas (LEAs) are considered \citep{afte-article1992}. Bullet lands are depressed areas between the  grooves made by the rifling action of the barrel. Compared to toolmarks made by screwdrivers striation marks on bullets are typically much smaller, both in length and in width. Bullets also have a curved cross-sectional topography.

@@

<!--Bullet matching methods are usually based on these associated signatures.-->
In same-source comparisons this curvature is usually removed using some form of Gaussian filter \citep{ma2004} or non-parametric smoothing \citep{aoas}. 
\citet{chu2013} use an automatic method for counting consecutive matching striae (CMS). The authors report an error rate of 52% for known same-source land comparisons to be (incorrectly) identified as different-source (false negative) and zero false positives for known different-source lands. \citet{ma2004} and \citet{vorburger2011} discuss CCF (cross-correlation function) and its discriminating power and applicability for same-source analyses of bullets, but do not provide any error rates in their discussion. \citet{aoas} use multiple features, such as CCF, CMS, D (distance measure), etc. in a random forest based method and compare every land against every other land of  digitized versions of Hamby 252 and Hamby 44 \citep{hamby} published on the NIST Ballistics Database \citep{nist}. The authors report an out-of-bag overall error rate of 0.46%, comprised of an error rate of 30.05% of  same-source pairs that were not identified and an error rate of 0.026% of different-source pairs that were incorrectly identified as same-source.

@@

\hh{What are we investigating in this paper, and what is the structure for the paper?}

@@

In a pre-processing step  the two markings are smoothed using a LOWESS \citep{lowess} with coarseness parameter $c$. 
 Originally, this smoothing is intended to remove drift and (sub)class characteristics from individual markings, however, in the setting of matching bullet striae, we can also make use of this mechanism to separate bullet curvature in profiles from signatures before matching signatures.  \autoref{fig:sigs-profiles} shows an example of a bullet land profile (left) and the corresponding signature (right).

@@

\autoref{fig:eer} shows a comparison of the performance of the two methods CS1 and CS2 with respect to EER (equal error rate) and AUC (area under the curve) corresponding to the ROC curves shown in \autoref{fig:roc}. Equal error rates are reduced using method CS2, while area under the curve significantly increases (at a significance level $\alpha$ of 0.05) compared to method CS1.

@@

cvcs %>% filter(!wo %in% c(130, 140)) %>%
  ggplot(aes(d = as.numeric(match), m = 1-p_value)) + 
  geom_roc(labels=FALSE, n.cuts=500, pointsize=0, size=0.6, 
           aes(colour = method.wo)) +
  facet_wrap(~wv, labeller = label_bquote(w[v] == .(wv))) +
  scale_alpha_discrete(range=c(0.1, 0.6)) +
  scale_colour_manual(expression("Method /\nOptimization\nwindow size"~w[o]), values=cols) +
  theme_bw() +
  xlab("False Positives Rate (1 - Specificity)") +
  ylab("True Positives Rate (Sensitivity)") +
  guides(
   colour = guide_legend(ncol=2)
  ) +
  geom_abline(slope=-1, intercept = 1, linetype = 3, 
              colour="grey30", size = 0.25)

@@

\begin{table}[hbtp]
\centering
\begin{adjustbox}{max width=\textwidth}
\begin{tabular}{rrrlrrl}
  \hline
  & \multicolumn{3}{l}{\bf CS1} & \multicolumn{3}{l}{\bf CS2} \\
$w_o$ & \bf FPR & \bf FNR & \bf AUC (95\% C.I.) & \bf FPR & \bf FNR & \bf AUC (95\% C.I.) \\ 
  \hline
90 & 0.068 & 0.330 & 0.850 (0.835, 0.865) & 0.075 & 0.242 & 0.877 (0.863, 0.890) \\ 
  120 & 0.065 & 0.309 & 0.864 (0.850, 0.878) & 0.072 & 0.236 & 0.890 (0.877, 0.903) \\ 
  130 & 0.065 & 0.317 & 0.863 (0.850, 0.877) & 0.072 & 0.224 & 0.898 (0.886, 0.910) \\ 
  140 & 0.065 & 0.304 & 0.873 (0.859, 0.886) & 0.071 & 0.229 & 0.902 (0.891, 0.914) \\ 
  150 & 0.067 & 0.326 & 0.863 (0.850, 0.877) & 0.072 & 0.225 & 0.904 (0.892, 0.916) \\ 
  180 & 0.061 & 0.326 & 0.877 (0.864, 0.890) & 0.070 & 0.229 & 0.907 (0.896, 0.919) \\ 
  210 & 0.064 & 0.342 & 0.865 (0.851, 0.878) & 0.066 & 0.231 & 0.906 (0.895, 0.918) \\ 
   \hline
\end{tabular}
\end{adjustbox}
\caption{\label{tab:aucs} Overview of results as shown in Figures \ref{fig:roc} and \ref{fig:eer}. FPR is the observed rate of false positives (for a fixed level $\alpha = 0.05$), FNR is the false negative rate. Area under the curve (AUC) is shown with confidence intervals. }
\end{table}

@@

The high-resolution scans of Hamby set 44 are capturing images at a resolution of $0.645 \mu m$ per pixel. On average, land engraved areas are 3000 pixels in length. A coarseness of $c = 0.125$ seemed to be sufficient in removing any bullet curvature. Both methods have a failed test rate of less than 0.006, indicating, again, that the larger number of pixels alleviates the problem of test failures.
\autoref{fig:h44} shows the resulting EER and AUC for methods CS1 and CS2 based on two sizes of validation windows ($w_v \in 75, 125$) and optimization window sizes around 300 pixels (10 percent of the average length). Both methods show an increase in performance around $w_o = 300$ pixels. CS2 out-performs CS1 in all scenarios, but the difference is not significant (using DeLong's confidence intervals). Interestingly, the overall performance of both CS1 and CS2 is a lot lower for the high resolution version of Hamby-44 than for the lower resolution scans. The area under the curve overall is significantly lower for the high resolution scans than for the previous set of scans. Partly, this might be due to the particular choice of the parameters, partly the higher-resolution scans might be picking up on real differences between the lands that the lower-resolution scans fail to detect.

@@

In assessing the suitability of the (deterministic) Chumbley Score for matching striae on bullet lands we have gained valuable insights into the process: method CS1 as proposed by \citet{hadler} has a strong dependency on the specific choice of parameters; the defaults suggested by \citet{hadler} for screw drivers are not directly applicable  for the smaller bullet lands. The coarseness parameter in particular has a strong impact on the performance of the test. However, we were able to suggest some heuristics based on the assumption that once sub-class characteristics are removed, optimal locations are distributed uniformly across the profile. For bullet lands we found  a coarseness value of $c = 0.15$ to be suitable for the low-resolution scans from NIST and a value of $c=0.125$ suitable for the higher-resolution scans from CSAFE.
Sizes for optimization windows $w_o$ were based on cross-validation to minimize overall type 2 error rates. Unfortunately, this kind of assessment is only feasible in the setting of a large study, such as the one we presented, and does not transfer immediately to case work, where a forensic examiner would only deal with a few identifications. Parameter settings for different studies should be investigated and reported to allow for a further refinement.

Method CS1 proposed by \citet{hadler} has a minimal type 2 error rate of 0.272 for an optimized window size of 140 pixels -- which is considerably higher than the error rates achieved on matching toolmarks, but is similar to other single-feature methods proposed for bullet matching. 
Unfortunately, method CS1 also has a high rate of failed tests -- situations, in which the algorithm does not provide a result, due to the way different-shift pairs are constructed. 
 Algorithm CS2 is introduced here as a remedy for failed tests by introducing an alternate version of choosing different-shift pairs. 
The algorithm CS2 introduced here is constructed in a way that achieves on average a ten-fold reduction in the number of failures. While reducing the failure rate, the algorithm also shows an increase in the power of the test. Type II error rates of CS2 reach a minimum of 0.217 for an optimized window size of 130 pixels. This increase in power of CS2 over CS1 should also apply to previous studies on toolmarks. It would be interesting to see these results using the adjusted algorithm. Unfortunately, none of the studies have made the data publicly accessible.

While significantly reduced over CS1, CS2 still has type 2 error rates on bullet lands that are  higher than the error rates achieved on the --much larger-- toolmarks.
Applying these methods to the high-resolution scans provided by CSAFE shows that better scanning methodology does not guarantee a better matching performance.

However, bullets usually have multiple lands -- in the case of Ruger P85s as used in the Hamby study, there are six lands for each bullet. We might be able to get more power out of the test by adapting CS2 to work in a bullet-to-bullet comparison.

@@

\caption{\label{same-shift-failure}Sketch of same-shift pairings. When the lag between optimal reference points is too large to accommodate a validation window in both signatures, both CS1 and CS2 fail.}
\end{figure}

### 2018-05-23 doc/paper.tex
conclusion + stab at abstract

  \author{
        Ganesh Krishnan \thanks{The authors gratefully acknowledge \ldots{}} \\
    Department of Statistics, Iowa State University\\
     and \\     Heike Hofmann \\
    Department of Statistics and CSAFE, Iowa State University\\
      }
  \maketitle
} \fi

@@

\bigskip
\begin{abstract}

\end{abstract}

@@

Analysis of these digitized markings require the use of statistical
methods which can quantify the scientific mechanism of comparing
markings and serve as basis for an error rate calcluation.
\citet{manytoolmarks2} define a relative distance metric and use it as
similarity measure between two toolmarks. \citet{manytoolmarks1} extract
many small segments in the markings of two toolmarks and compare
similiarity using a maximum pearson correlation coefficient. The
Chumbley scoring method, first introduced by \citet{chumbley}, uses a
similar but more extensive framework based on a Mann-Whitney U test of
the resulting correlation coefficients. This approach is
non-deterministic, because segments are chosen randomly. \citep{hadler}
make the score deterministic for each pair of toolmarks by choosing
segments for comparison systematically. This approach also ensures
independence between segements of striae. In this paper, we are
investigating the applicability of the Chumbley scoring method by
\citet{hadler} to assess striation marks on bullet lands for same-source
identification.

@@

In same-source comparisons this curvature is usually removed using some
form of Gaussian filter \citep{ma2004} or non-parametric smoothing
\citep{aoas}. \citet{chu2013} use an automatic method for counting
consecutive matching striae (CMS). The authors report an error rate of
52\% for known same-source land comparisons to be (incorrectly)
identified as different-source (false negative) and zero false positives
for known different-source lands. \citet{ma2004} and
\citet{vorburger2011} discuss CCF (cross-correlation function) and its
discriminating power and applicability for same-source analyses of
bullets, but do not provide any error rates in their discussion.
\citet{aoas} use multiple features, such as CCF, CMS, D (distance
measure), etc. in a random forest based method and compare every land
against every other land of digitised versions of Hamby 252 and Hamby 44
\citep{hamby} published on the NIST Ballistics Database \citep{nist}.
The authors report an out-of-bag overall error rate of 0.46\%, comprised
of an error rate of 30.05\% of same-source pairs that were not
identified and an error rate of 0.026\% of different-source pairs that
were incorrectly identified as same-source.

@@

In a pre-processing step the two markings are smoothed using a lowess
\citep{lowess} with coarseness parameter \(c\). Originally, this
smoothing is intended to remove drift and (sub)class characteristics
from individual markings, however, in the setting of matching bullet
striae, we can also make use of this mechanism to separate bullet
curvature in profiles from signatures before matching signatures.
\autoref{fig:sigs-profiles} shows an example of a bullet land profile
(left) and the corresponding signature (right).

@@

\autoref{fig:eer} shows a comparison of the performance of the two
methods CS1 and CS2 with respect to EER (eaual error rate) and AUC (area
under the curve) corresponding to the ROC curves shown in
\autoref{fig:roc}. Equal error rates are reduced using method CS2, while
area under the curve significantly increases (at a significance level
\(\alpha\) of 0.05) compared to method CS1.

@@

The high-resolution scans of Hamby set 44 are capturing images at a
resolution of \(0.645 \mu m\) per pixel. On average, land engraved areas
are 3000 pixels in length. A coarseness of \(c = 0.15\) was found to be
sufficient in removing any bullet curvature. \autoref{fig:h44} shows the
resulting EER and AUC for methods CS1 and CS2 based on two sizes of
validation windows (\(w_v \in 75, 125\)) and optimization window sizes
around 300 pixels (10 percent of the average length). Interestingly, the
overall performance of both CS1 and CS2 is a lot lower for the high
resolution version of Hamby-44 than for the lower resolution scans. The
area under the curve overall is significantly lower for the high
resolution scans than for the previous set of scans. Partly, this might
be due to the particular choice of the parameters.

@@

{\textcolor{orange}{
We started out trying to assess the suitability of the Chumbley Score's for matching striae on bullet lands. Besides using recommended defaults, we also proposed ways to optimize parameters: based on the assumption that once sub-class characteristics are removed, optimal locations are distributed uniformly across the profile, we found  for bullets the smaller coarseness value of $c = 0.15$ to be suitable. 
Optimized window sizes were found based on cross-validation to minimize type 2 error rates. 
Method CS1 proposed by \citet{hadler} has a minimal type 2 error rate of XXX for an optimized window size of YYY - which is considerably higher than the error rates achieved on matching toolmarks. 
Unfortunately, method CS1 also has a high rate of failed tests -- situations, in which the algorithm does not provide a result, due to the way different-shift pairs are constructed. 
XXX
The algorithm CS2 introduced here is constructed in a way that reduces the number of failures by a magnitude of about XXX. While reducing the failure rate, the algorithm also shows an increase in the power of the test. Type II error rates reach a minimum of XXX for an optimized window size of YYY. While significantly reduced over CS1, CS2 still has type 2 error rates on bullet lands that are  higher than the error rates achieved on the -much larger- toolmarks.
However, bullets usually have multiple lands - in the case of Ruger P85s as used in the Hamby study, there are six lands for each bullet. We might be able to get more power out of the test by adapting CS2 to deal work in a bullet-to-bullet comparison.
}}

{\textcolor{orange}{
Fix in CS2 should also improve power for matching toolmarks. 
}}

>>>>

  \author{
        Ganesh Krishnan \thanks{The authors gratefully acknowledge funding from CSAFE through
cooperative agreement \# 70NANB15H176 between NIST and Iowa State
University.} \\
    Department of Statistics, Iowa State University\\
     and \\     Heike Hofmann \\
    Department of Statistics and CSAFE, Iowa State University\\
      }
  \maketitle
} \fi

@@

\bigskip
\begin{abstract}
Same-source analyses are a major part of a Forensic Toolmark Examiner's
job. Technological advances, such as profilometers and confocal
microscopy allow to measure 3D surfaces in a high-resolution digitized
form. Methods for matching marks for a variety of tools have been
studied in the literature. Analysis of these digitized markings require
the use of statistical methods which can quantify the scientific
mechanism of comparing markings and serve as basis for an error rate
calculation. Striation marks on bullets are made by impurities in the
barrel. The Chumbley score provides us with another approach in the
same-source assessment of bullet striation marks.
\end{abstract}

@@

Analysis of these digitized markings require the use of statistical
methods which can quantify the scientific mechanism of comparing
markings and serve as basis for an error rate calculation.
\citet{manytoolmarks2} define a relative distance metric and use it as
similarity measure between two toolmarks. \citet{manytoolmarks1} extract
many small segments in the markings of two toolmarks and compare
similarity using a maximum Pearson correlation coefficient. The Chumbley
scoring method, first introduced by \citet{chumbley}, uses a similar but
more extensive framework based on a Mann-Whitney U test of the resulting
correlation coefficients. This approach is non-deterministic, because
segments are chosen randomly. \citep{hadler} make the score
deterministic for each pair of toolmarks by choosing segments for
comparison systematically. This approach also ensures independence
between segments of striae. In this paper, we are investigating the
applicability of the Chumbley scoring method by \citet{hadler} to assess
striation marks on bullet lands for same-source identification.

@@

In same-source comparisons this curvature is usually removed using some
form of Gaussian filter \citep{ma2004} or non-parametric smoothing
\citep{aoas}. \citet{chu2013} use an automatic method for counting
consecutive matching striae (CMS). The authors report an error rate of
52\% for known same-source land comparisons to be (incorrectly)
identified as different-source (false negative) and zero false positives
for known different-source lands. \citet{ma2004} and
\citet{vorburger2011} discuss CCF (cross-correlation function) and its
discriminating power and applicability for same-source analyses of
bullets, but do not provide any error rates in their discussion.
\citet{aoas} use multiple features, such as CCF, CMS, D (distance
measure), etc. in a random forest based method and compare every land
against every other land of digitized versions of Hamby 252 and Hamby 44
\citep{hamby} published on the NIST Ballistics Database \citep{nist}.
The authors report an out-of-bag overall error rate of 0.46\%, comprised
of an error rate of 30.05\% of same-source pairs that were not
identified and an error rate of 0.026\% of different-source pairs that
were incorrectly identified as same-source.

@@

In a pre-processing step the two markings are smoothed using a LOWESS
\citep{lowess} with coarseness parameter \(c\). Originally, this
smoothing is intended to remove drift and (sub)class characteristics
from individual markings, however, in the setting of matching bullet
striae, we can also make use of this mechanism to separate bullet
curvature in profiles from signatures before matching signatures.
\autoref{fig:sigs-profiles} shows an example of a bullet land profile
(left) and the corresponding signature (right).

@@

\autoref{fig:eer} shows a comparison of the performance of the two
methods CS1 and CS2 with respect to EER (equal error rate) and AUC (area
under the curve) corresponding to the ROC curves shown in
\autoref{fig:roc}. Equal error rates are reduced using method CS2, while
area under the curve significantly increases (at a significance level
\(\alpha\) of 0.05) compared to method CS1.

@@

The high-resolution scans of Hamby set 44 are capturing images at a
resolution of \(0.645 \mu m\) per pixel. On average, land engraved areas
are 3000 pixels in length. A coarseness of \(c = 0.125\) seemed to be
sufficient in removing any bullet curvature. Both methods have a failed
test rate of less than 0.006, indicating, again, that the larger number
of pixels alleviates the problem of test failures. \autoref{fig:h44}
shows the resulting EER and AUC for methods CS1 and CS2 based on two
sizes of validation windows (\(w_v \in 75, 125\)) and optimization
window sizes around 300 pixels (10 percent of the average length). Both
methods show an increase in performance around \(w_o = 300\) pixels. CS2
out-performs CS1 in all scenarios, but the difference is not significant
(using DeLong's confidence intervals). Interestingly, the overall
performance of both CS1 and CS2 is a lot lower for the high resolution
version of Hamby-44 than for the lower resolution scans. The area under
the curve overall is significantly lower for the high resolution scans
than for the previous set of scans. Partly, this might be due to the
particular choice of the parameters, partly the higher-resolution scans
might be picking up on real differences between the lands that the
lower-resolution scans fail to detect.

@@

In assessing the suitability of the (deterministic) Chumbley Score for
matching striae on bullet lands we have gained valuable insights into
the process: method CS1 as proposed by \citet{hadler} has a strong
dependency on the specific choice of parameters; the defaults suggested
by \citet{hadler} for screw drivers are not directly applicable for the
smaller bullet lands. The coarseness parameter in particular has a
strong impact on the performance of the test. However, we were able to
suggest some heuristics based on the assumption that once sub-class
characteristics are removed, optimal locations are distributed uniformly
across the profile. For bullet lands we found a coarseness value of
\(c = 0.15\) to be suitable for the low-resolution scans from NIST and a
value of \(c=0.125\) suitable for the higher-resolution scans from
CSAFE. Sizes for optimization windows \(w_o\) were based on
cross-validation to minimize overall type 2 error rates. Unfortunately,
this kind of assessment is only feasible in the setting of a large
study, such as the one we presented, and does not transfer immediately
to case work, where a forensic examiner would only deal with a few
identifications. Parameter settings for different studies should be
investigated and reported to allow for a further refinement.

Method CS1 proposed by \citet{hadler} has a minimal type 2 error rate of
0.272 for an optimized window size of 140 pixels -- which is
considerably higher than the error rates achieved on matching toolmarks.
Unfortunately, method CS1 also has a high rate of failed tests --
situations, in which the algorithm does not provide a result, due to the
way different-shift pairs are constructed. Algorithm CS2 is introduced
here as a remedy for failed tests by introducing an alternate version of
choosing different-shift pairs. The algorithm CS2 introduced here is
constructed in a way that achieves on average a ten-fold reduction in
the number of failures. While reducing the failure rate, the algorithm
also shows an increase in the power of the test. Type II error rates of
CS2 reach a minimum of 0.217 for an optimized window size of 130 pixels.
This increase in power of CS2 over CS1 should also apply to previous
studies on toolmarks. It would be interesting to see these results using
the adjusted algorithm. Unfortunately, none of the studies have made the
data publicly accessible.

While significantly reduced over CS1, CS2 still has type 2 error rates
on bullet lands that are higher than the error rates achieved on the
--much larger-- toolmarks. Applying these methods to the high-resolution
scans provided by CSAFE shows that better scanning methodology does not
guarantee a better matching performance.

However, bullets usually have multiple lands -- in the case of Ruger
P85s as used in the Hamby study, there are six lands for each bullet. We
might be able to get more power out of the test by adapting CS2 to work
in a bullet-to-bullet comparison.
